# Gambia Malaria Data

The `gambia` dataset from the R package geoR contains observations of $n = 2035$ Gambian children. The eight variables measured are:

* `x` the x-coordinate of the village (Universal Transverse Mercator - similar to latitude and longitude)
* `y` the y-coordinate of the village (UTM)
* `pos` presence (1) or absence (0) of malaria in a blood sample taken from the child
* `age` age of the child, in days
* `netuse` indicator variable denoting whether (1) or not (0) the child regularly sleeps under a bed-net
* `treated` indicator variable denoting whether (1) or not (0) the bed-net is treated (coded 0 if `netuse` = 0)
* `green` satellite-derived measure of the green-ness of vegetation in the immediate vicinity of the village (arbitrary units)
* `phc` indicator variable denoting the presence (1) or absence (0) of a health center in the village

## Logistic regression

Consider response $Y \in \{0, 1\}$ modelled as $Y \sim \text{Bern}(q)$ and covariates $x \in \mathbb{R}^p$ with

$$
\log\left(\frac{q(x)}{1-q(x)}\right) = \beta_0 + \beta^{T} x,
$$
where $\beta \in \mathbb{R}^p$. Then

$$
q(x) = \frac{\exp \left(\beta_0 + \beta^{T} x\right)}{1+ \exp\left(\beta_0 + \beta^{T} x\right)}
$$
```{r}
# Classify to 1 with probability
q <- function(x, b) {
  exp(b %*% x) / (1 + exp(b %*% x)) 
}
```

Observe labelled data $\mathcal{D} = \{(x_1, y_1), \ldots, (x_n, y_n)\}$. The likelihood function is

$$
\mathcal{L}(\beta_{0}, \beta) = \prod_{i=1}^{n} q\left(x_i\right)^{y_i}\left(1-q(x_i)\right)^{1-y_i}
$$
Place Gaussian priors on $\beta$ and $\beta_0$ such that

$$
\beta_0 \sim \mathcal{N}(\mu_0, \sigma_0^2), \;
\beta \sim \mathcal{N}_p(\mu, \text{diag}(\sigma_1^2, \ldots, \sigma_p^2))
$$
Then the posterior is proportional to

$$
p(\beta_0, \beta | y_1, \ldots, y_n) \propto \prod_{j=0}^p\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right) \prod_{i=1}^{n} q\left(x_i\right)^{y_i}\left(1-q\left(x_i\right)\right)^{1-y_i}.
$$
Taking the logarithm gives

$$
\log p(\beta_0, \beta | y_1, \ldots, y_n) \propto \sum_{j=0}^p \frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2 +  \sum_{i=1}^{n} \{ {y_i}\log q\left(x_i\right) + (1 - y_i) \log \left(1-q\left(x_i\right)\right)\}.
$$
The log-likelihood can be rewritten as

\begin{align*}
\sum_{i=1}^{n} \{ {y_i}\log q\left(x_i\right) + (1 - y_i) \log \left(1-q\left(x_i\right)\right) \}
&= \sum_{i=1}^{n} \{ {y_i}\log \left(\frac{q\left(x_i\right)}{1-q\left(x_i\right)}\right) + \log \left(1-q\left(x_i\right)\right) \} \\
&= \sum_{i=1}^{n} \{ {y_i}\log \left(\frac{q\left(x_i\right)}{1-q\left(x_i\right)}\right) + \log \left(1-q\left(x_i\right)\right) \} \\
&= \sum_{i=1}^{n} \{ {y_i}\left(\beta_0 + \beta^{T} x\right) - \log \left(1 + \exp(\beta_0 + \beta^{T} x)\right)\},
\end{align*}
so that the log-posterior is

$$
\log p(\beta_0, \beta | y_1, \ldots, y_n) \propto \sum_{j=0}^p \frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2 + \sum_{i=1}^{n} \{{y_i}\left(\beta_0 + \beta^{T} x\right) - \log \left(1 + \exp(\beta_0 + \beta^{T} x)\right)\}
$$

```{r}
# (proportional to) log posterior in the indep normals prior case
logpost <- function(b, X, mu, sigma) {
  logprior <- sum((b - mu)^2 / 2*sigma)
  nu <- apply(X, 1, function(x) b %*% x) # Vector of linear predictors
  loglike <- sum(nu[Y == 1]) + sum(-log(1 + exp(nu)))
  logprior + loglike
}
```

## Full and submodels

Firstly, the full model $\mathcal{M}$ is the logistic regression of response `pos` on the other variables including an intercept term but excluding the co-ordinates `x` and `y`.
$$
\log\left(\frac{q(x)}{1-q(x)}\right) = \eta
$$

$$
\eta = \beta_0 + \beta_1 \cdot \texttt{age} + \beta_2 \cdot \texttt{netuse} + \beta_3 \cdot \texttt{treated} + \beta_4 \cdot \texttt{green} + \beta_5 \cdot \texttt{phc}
$$

Define submodels $\mathcal{M}_1$ and $\mathcal{M}_2$ with linear predictors 
$$
\eta_1 = \beta_{0,1} + \beta_1 \cdot \texttt{age} +  \beta_4 \cdot \texttt{green} + \beta_5 \cdot \texttt{phc},
$$

and
$$
\eta_2 = \beta_{0,2} + \beta_2 \cdot \texttt{netuse} + \beta_3 \cdot \texttt{treated} + \beta_5 \cdot \texttt{phc}.
$$

|   | $\text{Intercept}$ | $\beta_1$ | $\beta_2$ | $\beta_3$ | $\beta_4$ | $\beta_5$ |
|---|:----:|----|----|----|----|----|
| Submodel 1 | $\checkmark$  | $\checkmark$  |   |   | $\checkmark$  | $\checkmark$  |
| Submodel 2 | $\checkmark$  |   | $\checkmark$  | $\checkmark$  |   | $\checkmark$  |

The link parameter is $\phi = \beta_5$ and model specific parameters are $\psi_1 = (\beta_{0,1}, \beta_1, \beta_4)$ and $\psi_2 = (\beta_{0,2}, \beta_2, \beta_3)$. Both submodels have the same observable random variables $Y_1 = Y_2 = Y$, the response variable `pos`.

Define $q_k$ for $k = 1, 2$ by
$$
q_k(x) = \frac{\exp \left(\eta_k\right)}{1+ \exp\left(\eta_k\right)}
$$

Take a normal prior as in Section 1 for $\beta_{1:5}$, and similarly take a normal prior for the intercepts
$$
\beta_{0, k} \sim \mathcal{N}(\mu_{0, k}, \sigma_{0, k}^2).
$$

Then the submodels have consistent prior marginals in the link parameter and Markov combination can be applied.

The joint distribution corresponding to submodel $\mathcal{M}_1$, as a function of the parameters, is proportional to the posterior, which itself is proportional to the prior times the likelihood

\begin{align*}
p_1(\phi, \psi_1, \mathbf{y}_1) &\propto p_1(\phi, \psi_1 | \mathbf{y}_1) \\
&= p_1(\beta_{0, 1}, \beta_1, \beta_4, \beta_5 | \mathbf{y}) \\
&\propto \underbrace{\exp\left(\frac{1}{2\sigma_{0,1}^2}(\beta_{0,1} - \mu_{0,1})^2\right) \prod_{j=1,4,5}\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right)}_{\text{Prior on } (\phi, \psi_1) = (\beta_{0,1}, \beta_1, \beta_4, \beta_5)} \times
\underbrace{\prod_{i=1}^{2035} q_1\left(x_i\right)^{y_i}\left(1-q_1\left(x_i\right)\right)^{1-y_i}}_{\text{Likelihood}}.
\end{align*}

Similarly for $\mathcal{M}_2$

\begin{align*}
p_2(\phi, \psi_2, \mathbf{y}_2) &\propto p_2(\phi, \psi_2 | \mathbf{y}_2) \\
&= p_2(\beta_{0, 2}, \beta_2, \beta_3, \beta_5 | \mathbf{y}) \\
&\propto \underbrace{\exp\left(\frac{1}{2\sigma_{0,2}^2}(\beta_{0,2} - \mu_{0,2})^2\right) \prod_{j=2,3,5}\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right)}_{\text{Prior on } (\phi, \psi_1) = (\beta_{0,2}, \beta_2, \beta_3, \beta_5)} \times 
\underbrace{\prod_{i=1}^{2035} q_2\left(x_i\right)^{y_i}\left(1-q_2\left(x_i\right)\right)^{1-y_i}}_{\text{Likelihood}}.
\end{align*}

Therefore the Markov combination in this case is

\begin{align*}
p_{\mathrm{comb}}\left(\phi, \psi_1, \psi_2, \mathbf{y}_1, \mathbf{y}_2\right) &= \frac{p_1\left(\phi, \psi_1, \mathbf{y}_1\right)p_2\left(\phi, \psi_2, \mathbf{y}_2\right)}{p(\phi)} \\
&\propto \frac{p_1(\beta_{0, 1}, \beta_1, \beta_4, \beta_5 | \mathbf{y}) p_2(\beta_{0, 2}, \beta_2, \beta_3, \beta_5 | \mathbf{y})}{p(\beta_5)} \\
&\propto \prod_{k=1}^2 \exp\left(\frac{1}{2\sigma_{0,k}^2}(\beta_{0,k} - \mu_{0,k})^2\right) \prod_{j=1}^5\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right) \\
&\times \prod_{i=1}^{2035} q_1\left(x_i\right)^{y_i}\left(1-q_1\left(x_i\right)\right)^{1-y_i} q_2\left(x_i\right)^{y_i}\left(1-q_2\left(x_i\right)\right)^{1-y_i} \tag{*}
\end{align*}

Informally $p_{\mathrm{comb}}$ contains the product of the priors on the full set of parameters $(\phi, \psi_1, \psi_2)$ with both likelihoods from $\mathcal{M}_1$ and $\mathcal{M}_2$. So this is almost like Bayesian inference for full model but with a different likelihood. It seems as if the information contained by the data $\mathcal{D}$ is being used more than once.

## Monte Carlo schemes

Want to sample from the target (*) using the methods described in [@goudie2019joining]. Continue to use (symmetric) normal proposals throughout.

### Metropolis-within-Gibbs

* Update first latent parameter $\psi_1$ by systematic scan Metropolis-within-Gibbs
* Update second latent parameter $\psi_2$ by systematic scan Metropolis-within-Gibbs
* Update link parameter $\phi$ by Metropolis-within-Gibbs

### Multi-stage Metropolis-within-Gibbs
