# Gambia Malaria Data

```{r echo=FALSE}
options(scipen = 999)
pacman::p_load(tidyverse, geoR, bayesplot)
```

The `gambia` dataset from the R package `geoR` contains observations of $n = 2035$ Gambian children. The eight measured variables are:

* `x` the x-coordinate of the village (Universal Transverse Mercator - similar to latitude and longitude)
* `y` the y-coordinate of the village (UTM)
* `pos` presence (1) or absence (0) of malaria in a blood sample taken from the child
* `age` age of the child, in days
* `netuse` indicator variable denoting whether (1) or not (0) the child regularly sleeps under a bed-net
* `treated` indicator variable denoting whether (1) or not (0) the bed-net is treated (coded 0 if `netuse` = 0)
* `green` satellite-derived measure of the green-ness of vegetation in the immediate vicinity of the village (arbitrary units)
* `phc` indicator variable denoting the presence (1) or absence (0) of a health center in the village

The aim is to fit models predicting `pos`, the presence of malaria, using the other variables (exclusing the positions `x` and `y`).

```{r}
data(gambia)
X <- as.matrix(gambia[, c(4:8)])
X <- cbind(intercept = 1, X) # Add intercept column to design matrix
Y <- gambia[, 3] # Response variable
n <- length(Y)
```

## Bayesian Logistic Regression

Consider response $Y \in \{0, 1\}$ modelled as $Y \sim \text{Bern}(q)$ and covariates $x \in \mathbb{R}^p$ with

$$
\log\left(\frac{q(x)}{1-q(x)}\right) = \beta_0 + \beta^{T} x,
$$
where $\beta \in \mathbb{R}$ and $\beta \in \mathbb{R}^p$. Then

$$
q(x) = \frac{\exp \left(\beta_0 + \beta^{T} x\right)}{1+ \exp\left(\beta_0 + \beta^{T} x\right)}
$$

```{r}
# Classify to 1 with probability
q <- function(x, b) {
  exp(b %*% x) / (1 + exp(b %*% x))
}
```

Observe labelled data $\mathcal{D} = \{(x_1, y_1), \ldots, (x_n, y_n)\}$. The likelihood function is

$$
\mathcal{L}(\beta_{0}, \beta) = \prod_{i=1}^{n} q\left(x_i\right)^{y_i}\left(1-q(x_i)\right)^{1-y_i}
$$
Place Gaussian priors on $\beta$ and $\beta_0$ such that

$$
\beta_0 \sim \mathcal{N}(\mu_0, \sigma_0^2), \; \beta \sim \mathcal{N}_p(\mu, \text{diag}(\sigma_1^2, \ldots, \sigma_p^2))
$$
Then the posterior is proportional to

$$
p(\beta_0, \beta | y_1, \ldots, y_n) \propto \prod_{j=0}^p\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right) \prod_{i=1}^{n} q\left(x_i\right)^{y_i}\left(1-q\left(x_i\right)\right)^{1-y_i}.
$$
Taking the logarithm gives

$$
\log p(\beta_0, \beta | y_1, \ldots, y_n) \propto \sum_{j=0}^p \frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2 +  \sum_{i=1}^{n} \{ {y_i}\log q\left(x_i\right) + (1 - y_i) \log \left(1-q\left(x_i\right)\right)\}.
$$
The log-likelihood can be rewritten as

\begin{align*}
\sum_{i=1}^{n} \{ {y_i}\log q\left(x_i\right) + (1 - y_i) \log \left(1-q\left(x_i\right)\right) \}
&= \sum_{i=1}^{n} \{ {y_i}\log \left(\frac{q\left(x_i\right)}{1-q\left(x_i\right)}\right) + \log \left(1-q\left(x_i\right)\right) \} \\
&= \sum_{i=1}^{n} \{ {y_i}\log \left(\frac{q\left(x_i\right)}{1-q\left(x_i\right)}\right) + \log \left(1-q\left(x_i\right)\right) \} \\
&= \sum_{i=1}^{n} \{ {y_i}\left(\beta_0 + \beta^{T} x\right) - \log \left(1 + \exp(\beta_0 + \beta^{T} x)\right)\},
\end{align*}

so that the log-posterior is
$$
\log p(\beta_0, \beta | y_1, \ldots, y_n) \propto \sum_{j=0}^p \frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2 + \sum_{i=1}^{n} \{{y_i}\left(\beta_0 + \beta^{T} x\right) - \log \left(1 + \exp(\beta_0 + \beta^{T} x)\right)\}
$$

```{r}
# (proportional to) log posterior in the indep normals prior case
logpost <- function(b, X, mu, sigma) {
  logprior <- sum((b - mu)^2 / 2*sigma)
  nu <- apply(X, 1, function(x) b %*% x) # Vector of linear predictors
  loglike <- sum(nu[Y == 1]) + sum(-log(1 + exp(nu)))
  logprior + loglike
}
```

## Full Model

The full model $\mathcal{M}$ is the logistic regression of response `pos` on the other variables including an intercept term but excluding the co-ordinates `x` and `y`.
$$
\log\left(\frac{q(x)}{1-q(x)}\right) = \eta
$$

$$
\eta = \beta_0 + \beta_1 \cdot \texttt{age} + \beta_2 \cdot \texttt{netuse} + \beta_3 \cdot \texttt{treated} + \beta_4 \cdot \texttt{green} + \beta_5 \cdot \texttt{phc}
$$

### Maximum-likelihood

Fitting a (frequentist) logisitic regression gives

```{r}
mle <- glm(Y ~ X - 1, family = "binomial") # Logistic regression using MLE
b <- mle$coefficients
summary(mle)
```

### Bayesian

Set independent normal priors with $\mu_j = 0$ and $\sigma_j = 0.1$ for all $j$, that is

$$
\beta_j \sim \mathcal{N}(0, 0.1) \; \forall j = 0, \ldots, 5
$$

Use Metropolis-within-Gibbs \@ref(mwg) to perform inference.

```{r}
# Function to update jth component of beta via RWMH step
mh <- function(b, X, mu, sigma, j, scale) {
  y <- b
  y[j] <- y[j] + rnorm(1, mean = 0, sd = scale[j])
  a <- exp(logpost(y, X, mu, sigma) - logpost(b, X, mu, sigma)) # Acceptance probability
  if(a > runif(1)) return(y) else return(b)
}

# Metropolis-within-Gibbs sampler (random scan by default)
mwg <- function(b0, X, mu, sigma, scale, nsim, random_scan = TRUE) {
  p <- length(b0)
  r <- array(NA, c(nsim, p)) # For the chain
  r[1, ] <- b0 # Init chain
  s <- array(0, c(3, p)) # For acceptance rates
  for(i in 2:nsim) {
    if(random_scan) {
      j <- sample(1:p, 1) # Random scan
    } else {
      j <- (i %% p) + 1 # Systematic scan
    }
    s[1, j] <- s[1, j] + 1 # Update pick count
    r[i, ] <- mh(r[i-1, ], X, mu, sigma, j, scale)
    if(!identical(r[i, ], r[i-1, ])) s[2, j] <- s[2, j] + 1 # Update accept count
  }
  r <- as.data.frame(r)
  names(r) <- sprintf("b%d", 0:(p-1))
  s[3, ] <- s[2, ] / s[1, ] # Acceptance rates
  return(list("chain" = r, "accept" = s))
}
```

Use `my_guess` for scaling (found by manual adjustment)

```{r}
my_guess <- c(0.175, 0.00025, 0.2, 0.3, 0.005, 0.2)
```

Run

```{r eval = FALSE}
full <- mwg(rep(0, 6), X, mu, sigma, scale = my_guess, nsim = 10^5)
```

```{r echo = FALSE}
full <- readRDS(file = "../output/full_model.Rds")
```

```{r warning=FALSE, message=FALSE}
# Want to do pairs plots but having trouble rendering
mcmc_trace(full$chain) # Traceplots
mcmc_hist(full$chain) # Histograms
full$accept
colMeans(full$chain); b # Pretty close to the ML estimates, so the code is probably correct
```

## Submodels

|   | $\text{Intercept}$ | $\beta_1$ | $\beta_2$ | $\beta_3$ | $\beta_4$ | $\beta_5$ |
|---|:----:|:--:|:--:|:--:|:--:|:--:|
| Submodel 1 | $\checkmark$  | $\checkmark$  |   |   | $\checkmark$  | $\checkmark$  |
| Submodel 2 | $\checkmark$  |   | $\checkmark$  | $\checkmark$  |   | $\checkmark$  |


The submodels $\mathcal{M}_1$ and $\mathcal{M}_2$ are defined by linear predictors 
$$
\eta_1 = \beta_{0} + \beta_1 \cdot \texttt{age} +  \beta_4 \cdot \texttt{green} + \beta_5 \cdot \texttt{phc},
$$

and
$$
\eta_2 = \beta_{0} + \beta_2 \cdot \texttt{netuse} + \beta_3 \cdot \texttt{treated} + \beta_5 \cdot \texttt{phc}.
$$

The link parameter is $\phi = (\beta_0, \beta_5)$ and model specific parameters are $\psi_1 = (\beta_1, \beta_4)$ and $\psi_2 = (\beta_2, \beta_3)$. Both submodels have the same observable random variables $Y_1 = Y_2$ the response variable `pos`.

Define $q_k$ for $k = 1, 2$ by
$$
q_k(x) = \frac{\exp \left(\eta_k\right)}{1+ \exp\left(\eta_k\right)}
$$

First we will run the submodels seperately.

### Maximum-likelihood

```{r}
X1 <- X[, c(1, 2, 5, 6)] # Design matrix 1: intercept, age, green and phc
p1 <- 4

X2 <- X[, c(1, 3, 4, 6)] # Design matrix 2: intercept, netuse, treated and phc
p2 <- 4

# GLM for model 1
mle1 <- glm(Y ~ X1 - 1, family = "binomial")
summary(mle1)

b1 <- mle1$coefficients

# GLM for model 2
mle2 <- glm(Y ~ X2 - 1, family = "binomial")
summary(mle2)

b2 <- mle2$coefficients
```

### Bayesian

Take the same normal prior as before for $\beta_{0:5}$.

```{r}
# Priors same as in the joint modelling to start with (so consistent in the link parameter)
mu1 <- rep(0, p1)
sigma1 <- rep(0.1, p1)

mu2 <- rep(0, p2)
sigma2 <- rep(0.1, p2)
```

```{r eval=FALSE}
model1 <- mwg(b0 = rep(0, 4), X = X1, mu = mu1, sigma = sigma1,
               scale = my_guess[c(1, 2, 5, 6)], nsim = 10^5)
names(model1$chain) <- c("b0", "b1", "b4", "b5") # Correct the naming

model2 <- mwg(b0 = rep(0, 4), X = X2, mu = mu2, sigma = sigma2,
               scale = my_guess[c(1, 3, 4, 6)], nsim = 10^5)
names(model2$chain) <- c("b0", "b2", "b3", "b5") # Correct the naming
```

```{r echo=FALSE}
model1 <- readRDS(file = "../output/model1.Rds")
model2 <- readRDS(file = "../output/model2.Rds")
```

```{r warning=FALSE, message=FALSE}
mcmc_trace(model1$chain) # Traceplots
mcmc_hist(model1$chain) # Histograms
model1$accept
colMeans(model1$chain); b1

mcmc_trace(model2$chain) # Traceplots
mcmc_hist(model2$chain) # Histograms
model2$accept
colMeans(model2$chain); b2
```

### Markov Combination

The submodels have consistent prior marginals in the link parameter and Markov combination can be applied.

The joint distribution corresponding to submodel $\mathcal{M}_1$, as a function of the parameters, is proportional to the posterior, which itself is proportional to the prior times the likelihood

\begin{align*}
p_1(\phi, \psi_1, \mathbf{y}_1) &\propto p_1(\phi, \psi_1 | \mathbf{y}_1) \\
&= p_1(\beta_0, \beta_1, \beta_4, \beta_5 | \mathbf{y}) \\
&\propto \underbrace{\prod_{j=0, 1,4,5}\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right)}_{\text{Prior on } (\phi, \psi_1) = (\beta_0, \beta_1, \beta_4, \beta_5)} \times
\underbrace{\prod_{i=1}^{2035} q_1\left(x_i\right)^{y_i}\left(1-q_1\left(x_i\right)\right)^{1-y_i}}_{\text{Likelihood}}.
\end{align*}

Similarly for $\mathcal{M}_2$
\begin{align*}
p_2(\phi, \psi_2, \mathbf{y}_2) &\propto p_2(\phi, \psi_2 | \mathbf{y}_2) \\
&= p_2(\beta_0, \beta_2, \beta_3, \beta_5 | \mathbf{y}) \\
&\propto \underbrace{\prod_{j=0, 2,3,5}\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right)}_{\text{Prior on } (\phi, \psi_1) = (\beta_0, \beta_2, \beta_3, \beta_5)} \times 
\underbrace{\prod_{i=1}^{2035} q_2\left(x_i\right)^{y_i}\left(1-q_2\left(x_i\right)\right)^{1-y_i}}_{\text{Likelihood}}.
\end{align*}

Therefore the Markov combination in this case is
\begin{align*}
p_{\mathrm{comb}}\left(\phi, \psi_1, \psi_2, \mathbf{y}_1, \mathbf{y}_2\right) &= \frac{p_1\left(\phi, \psi_1, \mathbf{y}_1\right)p_2\left(\phi, \psi_2, \mathbf{y}_2\right)}{p(\phi)} \\
&\propto \frac{p_1(\beta_0, \beta_1, \beta_4, \beta_5 | \mathbf{y}) p_2(\beta_0, \beta_2, \beta_3, \beta_5 | \mathbf{y})}{p(\beta_0, \beta_5)} \\
&\propto \prod_{j=0}^5\exp\left(\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2\right) \\
&\times \prod_{i=1}^{2035} q_1\left(x_i\right)^{y_i}\left(1-q_1\left(x_i\right)\right)^{1-y_i} q_2\left(x_i\right)^{y_i}\left(1-q_2\left(x_i\right)\right)^{1-y_i} (\#eq:target)
\end{align*}

Informally $p_{\mathrm{comb}}$ contains the product of the priors on the full set of parameters $(\phi, \psi_1, \psi_2)$ with both likelihoods from $\mathcal{M}_1$ and $\mathcal{M}_2$. So this is almost like Bayesian inference for full model but with a different likelihood. It seems as if the information contained by the data $\mathcal{D}$ is being used more than once.

Want to sample from \@ref(eq:target) using the methods described in @goudie2019joining. As before, it is convenient to work in terms of a logarithmic version

$$
\log p_{\mathrm{comb}}\left(\phi, \psi_1, \psi_2, \mathbf{y}_1, \mathbf{y}_2\right) = 
\sum_{j=0}^5\frac{1}{2\sigma_j^2}(\beta_j - \mu_j)^2 + \prod_{i=1}^{2035} q_1\left(x_i\right)^{y_i}\left(1-q_1\left(x_i\right)\right)^{1-y_i} q_2\left(x_i\right)^{y_i}\left(1-q_2\left(x_i\right)\right)^{1-y_i}
$$

#### Metropolis-within-Gibbs

* Update first latent parameter $\psi_1$ by systematic scan Metropolis-within-Gibbs
* Update second latent parameter $\psi_2$ by systematic scan Metropolis-within-Gibbs
* Update link parameter $\phi$ by Metropolis-within-Gibbs

#### Multi-stage Metropolis-within-Gibbs
